---
title: "Identifying Optimal Species Environments"
author:
  name: Joshua Ferrer-Lozano
  affiliation: Master of Environmental Data Science
date: 2025-11-28
categories: [R, MEDS, data-viz, Geospatial]
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
draft: false
---

# About

This blog explores the environmental suitability of oyster aquaculture along the U.S. West Coast. Using sea surface temperature (SST) data, bathymetry rasters, and Exclusive Economic Zone (EEZ) boundaries, we identify areas most favorable for oyster cultivation. By combining geospatial modeling with cartographic visualization, we highlight how environmental data science can inform sustainable aquaculture planning.

# Highlights of the Analysis

Raster Reclassification: Converted SST and depth rasters into binary suitability maps based on oyster thresholds.
Map Algebra: Combined multiple criteria (temperature × depth) to produce a composite suitability raster.
Zonal Statistics: Summed suitable areas within EEZ boundaries to quantify aquaculture potential by region.
Cartographic Visualization: Produced clear maps with legends, units, and orientation cues to communicate results effectively.

# Dataset Descriptions
- Sea Surface Temperature (2008–2012) NOAA Coral Reef Watch 5‑km SST data, averaged across five years and converted from Kelvin to Celsius. Accessed Nov. 2025.

- Bathymetry GEBCO gridded bathymetry dataset, reprojected and resampled to align with SST rasters. Accessed Nov. 2025.

- Exclusive Economic Zones (EEZs) Marine Regions shapefiles defining U.S. West Coast EEZ boundaries. Accessed Nov. 2025.

- Species Requirements Oyster depth and temperature preferences from SeaLifeBase. Accessed Nov. 2025.

# Findings

- Oyster aquaculture suitability is concentrated in specific EEZs along the West Coast.

- Combining SST and depth thresholds provides a robust method for identifying viable cultivation zones.

- Zonal statistics allow policymakers to compare aquaculture potential across regions.

# Conclusion

This analysis demonstrates how spatial data science can inform sustainable aquaculture planning. By integrating environmental thresholds with geospatial datasets, we can identify where oysters are most likely to thrive. Future work could expand this framework to other species — such as grapevines requiring specific elevations or trees adapted to certain climates — showcasing the versatility of raster reclassification and zonal statistics in environmental modeling.



# References

SeaLifeBase. (2025). Species depth and temperature requirements. https://www.sealifebase.ca/search.php [Accessed Nov. 2025]

NOAA Coral Reef Watch. (2025). 5‑km Sea Surface Temperature Anomaly Product. https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php [Accessed Nov. 2025]

GEBCO. (2025). Gridded Bathymetry Data. https://www.gebco.net/data-products/gridded-bathymetry-data#area [Accessed Nov. 2025]

Marine Regions. (2025). Exclusive Economic Zones (EEZs). https://www.marineregions.org/eez.php [Accessed Nov. 2025]

# Analysis 1: Mean Sea Surface Temperature

## Aquaculture Suitability Function

```{r}
# Import libraries for geospatial and data analysis
suppressPackageStartupMessages({
library(tidyverse)
library(stars)
library(terra)
library(sf)
library(tmap)
library(knitr)
library(kableExtra)
})
```


```{r}
# Function to map EEZ suitability for a species
map_species_suitability <- function(sst_min, sst_max, depth_min, depth_max, species_name) {
  
# Load in data
  regions_clean <- st_read("data/wc_regions_clean.shp", quiet = TRUE)
  depth <- rast("data/depth.tif")
  sst_stack <- c(
    rast("data/average_annual_sst_2008.tif"),
    rast("data/average_annual_sst_2009.tif"),
    rast("data/average_annual_sst_2010.tif"),
    rast("data/average_annual_sst_2011.tif"),
    rast("data/average_annual_sst_2012.tif")
  )
  
  # Process the SST and convert to celsius
  sst_mean <- mean(sst_stack)
  sst_celsius <- sst_mean - 273.15
  
  # Align depth to SST
  depth_proj <- project(depth, crs(sst_celsius))
  depth_crop <- crop(depth_proj, sst_celsius)
  depth_resample <- resample(depth_crop, sst_celsius, method = "near")
  
  # Create suitability classification matrixes
  sst_suit <- classify(sst_celsius,
                       rbind(c(-Inf, sst_min, 0),
                             c(sst_min, sst_max, 1),
                             c(sst_max, Inf, 0)))
  
  depth_suit <- classify(depth_resample,
                         rbind(c(-Inf, depth_min, 0),
                               c(depth_min, depth_max, 1),
                               c(depth_max, Inf, 0)))
  
  suitability <- sst_suit * depth_suit
  
  # Create EEZ zonal stats
  
  # Convert cleaned EEZ regions to vector
  eez_v <- vect(regions_clean)
  
  # Reproject the EEZ vector to match CRS of sst_celsius
  eez_v <- project(eez_v, crs(sst_celsius))
  
  # Add unique ID for each EEZ zone for later join
  eez_v$eez_id <- seq_len(nrow(eez_v))
  
  # Rasterize the EEZ polygon so each cell in raster is assigned an ID that is within
  eez_r <- rasterize(eez_v, sst_celsius, field = "eez_id")
  
  # Calculate the area of each raster in square kilometers
  cell_area <- cellSize(sst_celsius, unit = "km")
  
  # Multiply the suitability raster by cell area to get area of suitable habitat per cell
  area_suitable <- suitability * cell_area
  
  # Perform zonal math by getting the sum of the suitable area in each EEZ zone
  area_by_eez <- zonal(area_suitable, eez_r, fun = "sum", na.rm = TRUE)
  
  # Convert above results into a dataframe and rename column
  area_by_eez <- as.data.frame(area_by_eez) |> 
    rename(suitable_km2 = mean)
  
  # Convert EEZ vector back to an sf obejct and join stats by EEZ ID
  eez_sf_out <- st_as_sf(eez_v) |> 
    left_join(area_by_eez, by = "eez_id")
  
# Plot map
  tm_shape(eez_sf_out) +
    tm_polygons(fill = "suitable_km2",
                fill.scale = tm_scale(values = "viridis"),
                fill.legend = tm_legend(title = "Suitable area (km²)")) +
    tm_title(paste(species_name, "Suitability Across West Coast EEZs"),
             position = c("center", "top")) +
    tm_scalebar(position = c("left", "bottom")) +
    tm_compass(position = c("right", "top"))
}
```

```{r}
# Use the function for finding suitable area for Geoducks

# map_species_suitability(sst_min, sst_max, depth_min, depth_max, species_name)
map_species_suitability(sst_min = 0, sst_max = 80, depth_min = -100, depth_max = 0, species_name = "Geoduck")

```

# Analysis 1: Mean Sea Surface Temperature

## Import and Prepare the Data

```{r}
# Read in Exclusive Economic Zone (EEZ) shapefile
regions_clean <- st_read("data/wc_regions_clean.shp")

# Read in depth raster
depth <- rast("data/depth.tif")

# Read in Sea Surface Temperature rasters
sst_2008 <- rast("data/average_annual_sst_2008.tif")
sst_2009 <- rast("data/average_annual_sst_2009.tif")
sst_2010 <- rast("data/average_annual_sst_2010.tif")
sst_2011 <- rast("data/average_annual_sst_2011.tif")
sst_2012 <- rast("data/average_annual_sst_2012.tif")
```

## Process the Data

We assume WGS84 (EPSG:4326) as the common CRS. If CRS mismatches occur, analysis would be invalid.

```{r}
# Stack SST rasters into multi-layer Spatraster
sst_stack <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

```{r}
# Compute the mean of SST across years
sst_mean <- mean(sst_stack)
```

```{r}
# Convert Kelvin to Celsius
sst_celsius <- sst_mean - 273.15

# Replace NA values with 0
sst_celsius[is.na(sst_celsius)] <- 0

# Add warning if the range is wrong
if (max(values(sst_celsius), na.rm=TRUE) > 40) warning("SST unusually high — check conversion.")

```

```{r}
# Convert SST raster to stars object for a more robust plot
sst_stars <- st_as_stars(sst_celsius)

tm_shape(sst_stars) +
  tm_raster(fill.scale = tm_scale(values = "viridis"),
            fill.legend = tm_legend(title = "Temperature (C)")) +
  tm_title("Mean Sea Surface Temperature (°C, 2008–2012)", position = c("center", "top")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "bottom"))

```

## Reflection on Mean Sea Surface Temperature Raster

The mean sea surface temperature (SST) raster from the years 2008 - 2012 provides insight into the temperature variation along the west coast of the united states from California to Washington state. Because temperature data is continuous, a color scale was used to illustrate temperature variability. The above illustrate is not varied because bins of mean temperature across five years of sea surface temperature.

# Analysis 2: Combined Suitability Raster

```{r}
# Align depth raster to the SST raster

# Reproject CRS
depth_proj <- project(depth, crs(sst_celsius))  
# Crop to SST extent
depth_crop <- crop(depth_proj, sst_celsius)
# Resample resolution
depth_resample <- resample(depth_crop, sst_celsius, method = "near") 
```

Resampling with the nearest neighbor preserves categorical depth bins.

```{r}
# Verify alignments of CRS, resolution and extent
crs(sst_celsius) == crs(depth_resample)
if (!identical(res(sst_celsius), res(depth_resample))) stop("Resolution mismatch.")

```

```{r}
# Create matrix to reclassify SST suitability for oysters based on 11–30 C
sst_suit <- classify(sst_celsius,
                     rbind(c(-Inf, 11, 0), # Unsuitable temp below 11 C
                           c(11, 30, 1),   # Suitable temp at 11 - 30 C
                           c(30, Inf, 0))) # Unsuitable temp above 30 C

# Create matrix to reclassify depth suitability for oysters based on 0–70 m
depth_suit <- classify(depth_resample,
                       rbind(c(-Inf, -70, 0),  # Unsuitable elevation above sea level
                             c(-70, 0, 1),     # Suitable between -70 - 0 m sea level
                             c(0, Inf, 0)))    # Unsuitable elevation above sea level

# Test to verify our matrix of binary output
testthat::expect_true(all(unique(values(sst_suit)) %in% c(0,1)))
```

We assume depths are negative, so suitable oyster depths are −70 to 0 m. If positive, thresholds would be 0–70.

```{r}

# Combine suitability (map algebra: both must be 1)
suitability <- sst_suit * depth_suit

# Convert raster to dataframe
suit_df <- as.data.frame(suitability, xy = TRUE, na.rm = TRUE)

# Plot with ggplot
ggplot(suit_df) +
  geom_raster(aes(x = x, y = y, fill = factor(mean))) +
  scale_fill_manual(values = c("0" = "purple", "1" = "gold"),
                    name = "Oyster suitability",
                    labels = c("Unsuitable", "Suitable")) +
  labs(title = "Combined Suitability (Oysters)") +
  theme_minimal()

```

## Reflection on Combined Suitability Raster

The bathymetry raster illustrates ocean depth across the study area. With the given information, suitability classification was determined by depth preferences of oysters and the categories mapped "suitable" and "unsuitable" was based on binary outcomes of the suitable matrix sea depth. This binary map showcases environments suitable for oyster aquaculture.

# Analysis 3: Suitability by EEZ

```{r}
# Convert EEZ polygons to terra vector
eez_v <- vect(regions_clean)
# Align CRS
eez_v <- project(eez_v, crs(sst_celsius))

# Rasterize EEZ IDs to each polygon
eez_v$eez_id <- seq_len(nrow(eez_v))
eez_r <- rasterize(eez_v, sst_celsius, field = "eez_id")

# Calculate cell area in km^2
cell_area <- cellSize(sst_celsius, unit = "km")

# Multiply suitable area raster by cell to get suitable area per cell
area_suitable <- suitability * cell_area

# Zonal sum of suitable area per EEZ
area_by_eez <- zonal(area_suitable, eez_r, fun = "sum", na.rm = TRUE)

# Convert to data frame and rename columns correctly
area_by_eez <- as.data.frame(area_by_eez) |> 
  dplyr::rename(suitable_km2 = mean)
```

```{r}
head(arrange(area_by_eez, desc(suitable_km2)))
```

Summing suitable cell areas by EEZ allows us to rank zones by aquaculture potential

```{r}
# Join zonal statistics to EEZ polygons
eez_sf_out <- st_as_sf(eez_v) |> 
  left_join(area_by_eez, by = "eez_id")
```

```{r}
# Plot map of suitable area by EEZ
eez_sf <- st_as_sf(eez_v) |>  left_join(area_by_eez, by = "eez_id")

ggplot(eez_sf) +
  geom_sf(aes(fill = suitable_km2), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "C", name = "Suitable area (km^2)") +
  labs(title = "Oyster Suitability Across West Coast EEZs") +
  theme_minimal() +
  theme(
    plot.title.position = "plot",                # move title above plot panel
    plot.title = element_text(margin = margin(b = 10)) # add spacing below title
  )
```

## Reflection on Oyster Suitability Across West Coast EEZs

This map aggregates suitability into Exclusive Economic Zones (EEZs), showing total suitability across the zones and illustrates aquaculture potential. We see that based on the table below that EEZ ID magnitude of suitability from largest to smallest are EEZ IDs: 3,4,5,1 and 2 illustrating the southern west coast tend to be an optimal environment for Oyster aquaculture.

```{r}
# Create table of suitable areas by EEZ
area_by_eez %>%
  arrange(desc(suitable_km2)) %>%
  knitr::kable(digits = 2, caption = "Suitable Area by EEZ (km^2)") |> 
  kableExtra::kable_classic(full_width = FALSE, position = "center")
```
